Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-12-06 manuscript.qmd
start working on Jan's comments (during meeting)

  Results: The model accurately and precisely describes observed patterns of lymph node involvement with a compact set of interpretable parameters. Midline extension of the primary tumor is identified as the primary risk factor for contralateral involvement, with advanced T-stage and extensive ipsilateral involvement further increasing risk. The model highlights key hierarchical dependencies: occult disease in contralateral LNL III is highly unlikely if upstream LNL II is clinically negative, and in contralateral LNL IV, occult disease is exceedingly rare without LNL III involvement.

  Conclusions: This model offers an interpretable, probabilistic framework to inform personalized elective CTV-N volume reduction. For lateralized tumors that do not cross the midline, it suggests the contralateral neck may safely be excluded from elective irradiation. For tumors extending across the midline but with a clinically negative contralateral neck, elective irradiation could be limited to LNL II, reducing unnecessary exposure while maintaining oncological safety.
---

@@

```{python}
#| echo: false
#| label: tbl-data-overview
#| tbl-cap: Overview over the five datasets from four different institutions used to train and evaluate our model. Here, we briefly characterize the total number of OPSCC patients from the respective institution, their median age, what proportion received some form of neck dissection, the N0 portion of patients, what percentage presented with early T-category, and the prevalence of primary tumor midline extension. For a much more detailed look at the data, visit [lyprox.org](https://lyprox.org).
from typing import Literal

@@

Pathological involvement is available only for surgically treated patients and for the levels that were dissected. For non-surgical patients, involvement status is determined clinically, i.e. using imaging. To ensure uniformity in analysis, diagnostic information was synthesized into a consensus decision for each patient and LNL. This consensus reflects the most likely state of involvement and accounts for the sensitivity and specificity of various diagnostic modalities, as reported in the literature [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008].

@@

![Contralateral involvement stratified by T-category (top left panel), the number of metastatic LNLs ipsilaterally (top right panel), and whether the primary tumor extended over the mid-sagittal line or was clearly lateralized (bottom left panel). Additionally, in the bottom right, we show that even for lateralized tumors, T-category and ipsilateral involvement are correlated with contralateral metastasis by comparing the contralateral involvement prevalence for selected scenarios that vary in their T-category and ipsilateral involvement extent.](figures/fig_data_strat.svg){#fig-data-strat}

@@

The top right panel reveals a positive correlation between ipsilateral and contralateral metastases. Extensive ipsilateral involvement likely indicates a longer or faster disease progression. Additionally, it has been hypothesized that bulky ipsilateral nodal disease may reroute lymphatic drainage toward the contralateral side, potentially facilitating contralateral metastasis. <!-- citation needed? -->

### Interplay of Risk Factors

Midline extension, T-category, and ipsilateral involvement are interrelated risk factors for contralateral metastasis. For instance, {{< var data.late_with_midext_percent >}}% of advanced T-category tumors exhibit midline extension compared to {{< var data.early_with_midext_percent >}}% of early T-category tumors. While the higher fraction of midline extensions in advanced T-category patients partially accounts for the higher contralateral metastasis rates, T-category itself and ipsilateral involvement also play an additional role.

The bottom right panel of @fig-data-strat isolates patients with lateralized tumors that do not cross the midline, further demonstrating these relationships. Among early T-category patients with no ipsilateral nodal involvement (levels I-V), only {{< var data.early_lateral_ipsin0_cII_percent >}}% ({{< var data.early_lateral_ipsin0_cII_match >}} of {{< var data.early_lateral_ipsin0_cII_total >}} patients) show involvement in contralateral level II. This proportion increases to {{< var data.early_lateral_ipsiII_cII_percent >}}% ({{< var data.early_lateral_ipsiII_cII_match >}} of {{< var data.early_lateral_ipsiII_cII_total >}}) if ipsilateral level II is involved, to {{< var data.early_lateral_ipsiIIandIII_cII_percent >}}% ({{< var data.early_lateral_ipsiIIandIII_cII_match >}} of {{< var data.early_lateral_ipsiIIandIII_cII_total >}}) if ipsilateral levels II and III are involved, and further to {{< var data.late_lateral_ipsiIIandIII_cII_percent >}}% ({{< var data.late_lateral_ipsiIIandIII_cII_match >}} of {{< var data.late_lateral_ipsiIIandIII_cII_total >}}) for advanced T-category tumors with ipsilateral levels II and III involved.

@@

1. **Midline Extension:** Tumors extending across the mid-sagittal plane should result in a significantly higher probability of contralateral metastases.
2. **T-Category and Disease Progression:** Advanced T-category should correspond to an increased risk of nodal disease. This can be modeled using the expected time of diagnosis, as demonstrated in a prior hidden Markov model [@ludwig_hidden_2021].
3. **Ipsilateral Involvement as a Progression Marker:** The extent of ipsilateral involvement should provide additional information on the time elapsed between disease onset and diagnosis, supplementing the insights gained from T-category alone.

@@

This paper builds on the previously developed unilateral model for ipsilateral lymph node involvement presented in [@ludwig_modelling_2024]. Below, we briefly recap the unilateral model to introduce the notation required for extending the framework to a bilateral model, described in @sec-ext-to-contra. For further details on the ipsilateral model, refer to earlier publications [@ludwig_hidden_2021;@ludwig_modelling_2024].

@@

Here, $V$ is the number of LNLs the model considers. The values a LNL's may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is healthy or free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to some form of tumor presence (i.e., occult or macroscopic disease). In total, there are $2^V$ distinct possible lymphatic involvement patterns, which we enumerate from $\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}$ to $\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}$. Each LNL's state is observed via another binary random variable $Z_v$ that describes the clinical involvement of a LNL based on imaging: $Z_v = 0$ (`False`) indicates that the LNL $v$ is healthy based on clinical diagnosis, and $Z_v = 1$ (`True`) indicates that LNL $v$ was classified as involved. $X_v$ and $Z_v$ are connected through the sensitivity and specificity of the diagnositc modality.

@@

The transition matrix $\mathbf{A}$ is parameterized using a directed acyclic graph (DAG) that represents the underlying lymphatic network. Edges from the primary tumor to an LNL are associated with a probability $b_v$ for direct spread to LNL $v$ during one time step. Arcs from a LNL $v$ to a LNL $r$ are parameterized with the probability rate $t_{vr}$ for inter-nodal progression. In this paper, we build on the DAG shown in @fig-full-graph which was obtained by maximizing the model evidence as described in [@ludwig_modelling_2024].

@@

$$
P\left( \mathbf{X} \mid \text{T}x = \text{early} \right) = \sum_{t=0}^{t_\text{max}} P \left( \mathbf{X} \mid t \right) \cdot P(t \mid \text{early})
$$

@@

The likelihood function for a patient to present with a diagnosis $\boldsymbol{\zeta}_i$ a T-category $\text{T}x$ tumor can now be written as:

@@

With $\left[ \ldots \right]_i$ we denote the $i$-th component of the vector in the square brackets. Note that it is also possible to account for missing involvement information: If a diagnosis (like fine needle aspiration (FNA)) is only available for a subset of all LNLs, we can sum over all those possible complete observed states $\boldsymbol{\zeta}_j$ that match the provided diagnosis.

@@

<!-- This is displayed in @tbl-data-strat and shows how often the contralateral LNLs I, II, III, and IV were involved, given all possible combinations of midline extension, T-category, and ipsilateral LNL III involvement. Unsurprisingly, the prevalence for contralateral involvement is consistently higher when the tumor extends over the mid-sagittal line or is of later T-category. But it is also more frequent when the ipsilateral side shows more severe involvement, which is here shown via the surrogate LNL III. -->

@@

Thus, we extend the formalism in @sec-unilateral in such a way that the model's ipsi- and contralateral side evolve synchronously over time. To achieve that, we start by writing down the posterior distribution of involvement in analogy to @eq-uni-bayes-law, which is now a joint probability of an involvement $\mathbf{X}^\text{i}$ ipsilaterally *and* an involvement $\mathbf{X}^\text{c}$ contralaterally, given a diagnosis of the ipsilateral LNLs $\mathbf{Z}^\text{i}$ and of the contralateral ones $\mathbf{Z}^\text{c}$:

@@

The matrices $\boldsymbol{\Lambda}\text{i}$ and $\boldsymbol{\Lambda}\text{c}$ could, in principle, be parameterized with entirely separate parameters, allowing ipsilateral and contralateral spread rates to differ completely. However, we simplify the parameter space by sharing parameters between the two sides, based on the following three assumptions:

1. **Shared Graph Structure**: Both ipsilateral and contralateral spread are described by the same graph shown in @fig-full-graph.
2. **Symmetric Spread Among LNLs**: The spread among LNLs is assumed to be the same on both sides, reflecting the symmetric structure of the lymphatic system. Consequently, the spread rates between nodes should also be symmetric. This is formalized as:\
  $$
  \begin{aligned}
  b_v^\text{c} &\neq b_v^\text{i} \\
  t_{rv}^\text{c} &= t_{rv}^\text{i}
  \end{aligned}
  $$ {#eq-symmetries}
  for all $v \leq V$ and $r \in \operatorname{pa}(v)$.
3. **Asymmetric Spread from Tumor**: Tumor spread to the contralateral side varies depending on whether the tumor crosses the mid-sagittal plane. When the tumor extends over the midline, its spread rate to the contralateral LNLs should be larger than if the tumor were lateralized, but still smaller than the tumor's spread to the ipsilateral side. To model this, we introduce a mixing parameter $\alpha \in [0,1]$, representing the "degree of asymmetry" effectively mixing these two extremes:\
  $$
  b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
  $$ {#eq-mixing}
  This parameterization results in three sets of rates for tumor spread to the LNLs: (1) the spread to ipsilateral LNLs $b^\text{i}_v$, (2) the spread to contralateral LNLs as long as the tumor is lateralized $b_v^{\text{c},\epsilon=\texttt{False}}$, (3) the spread to contralateral LNLs when the tumor crosses the midline $b_v^{\text{c},\epsilon=\texttt{True}}$. Despite this additional complexity, these rates collectively account for only $2 \cdot 2^V + 1$ parameters due to their coupling via $\alpha$.

@@

To account for this, we model the tumor's extension over the mid-sagittal line as a binary random variable. A tumor starts as lateralized, with a finite probability $p_\epsilon$ at each time step of crossing the midline. The overall probabilities of a patient having a clearly lateralized tumor or one extending over the mid-sagittal line after $t$ time steps are then given by

@@

The case of midline extension is more complex: we already marginalize over the exact time step when the tumor grows over the mid-sagittal line. However, at the point of crossing, the contralateral transition matrix must switch to the increased spread rates, $b_v^{\text{c}, \epsilon=\texttt{True}}$, as defined by the linear mixing in @eq-mixing. To correctly perform this marginalization, we iteratively construct the joint distribution $P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)$.

@@

Using this, we can again write the joint of ipsi- and contralateral involvement - now also for the case of mid-sagittal extension - algebraically as before in @eq-bilateral-marginal-algebra:

@@

# Computational methods {#sec-methods}

@@

A "scenario" includes the patient's T-category $\text{T}x$ and whether the tumor extended over the mid-sagittal line, i.e. $\epsilon=\texttt{True}$ or $\epsilon=\texttt{False}$. An involvement pattern specifies all ipsi- and contralateral LNLs' status as "healthy", "involved", or "masked" (ignored).

@@

![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences, shown for the contralateral LNLs II (blue), III (orange), and IV (green). The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones. The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line. This figure illustrates the model's ability to describe the prevalence of involvement for different combinations of the risk factors T-category and midline extension.](figures/fig_model_prevalences_overall.svg){#fig-model-prevalences-overall}

@@

![Comparison of the computed and observed prevalences for scenarios that illustrate the model's capability of accounting for the correlation between ipsi- and contralateral involvement. We show three scenarios where we consider the joint involvement of contralateral LNL II together with different ipsilateral involvements: 1) the ipsilateral neck shows no involvement in green (LNLs I to V are healthy, LNL VII is unspecified because data on it is missing for some patients), 2) where ipsilateral LNL II is involved in orange (LNLs I, III, IV, and V are healthy), and 3) where ipsilateral LNLs II and III are involved in red (LNLs I, IV, and V are healthy). These three scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in the right column).](figures/fig_model_prevalences_with_ipsi.svg){#fig-model-prevalences-with-ipsi}

@@

```{python}
#| echo: false
#| label: tbl-data-strat
#| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal line, the T-category, and whether the ipsilateral LNL III was involved or healthy.
lnl_cols = get_lnl_cols("contra", lnls=["I", "II", "III", "IV"])
num_ipsi_inv = raw[get_lnl_cols("ipsi")].sum(axis="columns")

>>>>

  Results: The model accurately and precisely describes observed patterns of lymph node involvement with a compact set of interpretable parameters. Midline extension of the primary tumor is identified as the primary risk factor for contralateral involvement, with advanced T-stage and extensive ipsilateral involvement further increasing risk. Occult disease in contralateral LNL III is highly unlikely if upstream LNL II is clinically negative, and in contralateral LNL IV, occult disease is exceedingly rare without LNL III involvement.

  Conclusions: This model offers an interpretable, probabilistic framework to inform personalized elective CTV-N volume reduction. For lateralized tumors that do not cross the midline, it suggests the contralateral neck may safely be excluded from elective irradiation. For tumors extending across the midline but with a clinically negative contralateral neck, elective irradiation could be limited to LNL II, reducing unnecessary exposure of normal tissue while maintaining regional tumor control.
---

@@

```{python}
#| echo: false
#| label: tbl-data-overview
#| tbl-cap: Overview over the five datasets from four different institutions used to train and evaluate our model. Here, we briefly characterize the total number of OPSCC patients from the respective institution, their median age, what proportion received neck dissection, the N0 portion of patients, what percentage presented with early T-category, and the prevalence of primary tumor midline extension. For a much more detailed look at the data, visit [lyprox.org](https://lyprox.org).
from typing import Literal

@@

Pathological involvement is available only for surgically treated patients and for the levels that were dissected. For non-surgical patients, involvement status is determined clinically, i.e. using imaging. For this work, diagnostic information was synthesized into a consensus decision for each patient and LNL. This consensus reflects the most likely state of involvement and accounts for the sensitivity and specificity of various diagnostic modalities, as reported in the literature [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008].

@@

![Contralateral involvement stratified by T-category (top left panel), the number of metastatic LNLs ipsilaterally (top right panel), and whether the primary tumor extended over the mid-sagittal plane or was clearly lateralized (bottom left panel). In the bottom right panel, we consider lateralized tumors only, and compare the contralateral involvement prevalence for selected scenarios that vary in their T-category and ipsilateral involvement extent.](figures/fig_data_strat.svg){#fig-data-strat}

@@

The top right panel reveals a positive correlation between ipsilateral and contralateral metastases. Extensive ipsilateral involvement likely indicates a longer or faster disease progression. Additionally, it has been hypothesized that bulky ipsilateral nodal disease may reroute lymphatic drainage toward the contralateral side, potentially increasing the probability of contralateral metastasis. <!-- citation needed? -->

### Correlation of Risk Factors

Midline extension, T-category, and ipsilateral involvement are interrelated risk factors for contralateral metastasis. For instance, {{< var data.late_with_midext_percent >}}% of advanced T-category tumors exhibit midline extension compared to {{< var data.early_with_midext_percent >}}% of early T-category tumors. While the higher fraction of midline extensions in advanced T-category patients partially explains the higher contralateral metastasis rates, T-category itself and ipsilateral involvement also play an additional role.

The bottom right panel of @fig-data-strat considers only patients with lateralized tumors that do not cross the midline. Among early T-category patients with no ipsilateral nodal involvement (levels I-V), only {{< var data.early_lateral_ipsin0_cII_percent >}}% ({{< var data.early_lateral_ipsin0_cII_match >}} of {{< var data.early_lateral_ipsin0_cII_total >}} patients) show involvement in contralateral level II. This proportion increases to {{< var data.early_lateral_ipsiII_cII_percent >}}% ({{< var data.early_lateral_ipsiII_cII_match >}} of {{< var data.early_lateral_ipsiII_cII_total >}}) if ipsilateral level II is involved, to {{< var data.early_lateral_ipsiIIandIII_cII_percent >}}% ({{< var data.early_lateral_ipsiIIandIII_cII_match >}} of {{< var data.early_lateral_ipsiIIandIII_cII_total >}}) if ipsilateral levels II and III are involved, and further to {{< var data.late_lateral_ipsiIIandIII_cII_percent >}}% ({{< var data.late_lateral_ipsiIIandIII_cII_match >}} of {{< var data.late_lateral_ipsiIIandIII_cII_total >}}) for advanced T-category tumors with ipsilateral levels II and III involved.

@@

1. **Midline Extension:** Tumors extending across the mid-sagittal plane should result in a significantly higher probability of contralateral metastases.
2. **T-Category:** Advanced T-category should correspond to an increased risk of nodal disease. In the hidden Markov model this can be modeled using the expected time of diagnosis, as demonstrated previously for the ipsilateral model [@ludwig_hidden_2021].
3. **Ipsilateral Involvement:** The model should be able to capture the correlation between the extent of ipsi- and contralateral involvement. <!-- TODO: finish -->

@@

This paper builds on the previously developed unilateral model for ipsilateral lymph node involvement presented in [@ludwig_modelling_2024]. Below, we briefly recap the unilateral model to introduce the notation required for extending the framework to a bilateral model, described in @sec-ext-to-contra. For further details on the ipsilateral model, we refer to earlier publications [@ludwig_hidden_2021;@ludwig_modelling_2024].

@@

Here, $V$ is the number of LNLs the model considers. The values a LNL's may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is healthy or free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to the presence of clinically detected metastases (i.e., occult or macroscopic disease). In total, there are $2^V$ distinct possible lymphatic involvement patterns, which we enumerate from $\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}$ to $\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}$. Each LNL's state is observed via another binary random variable $Z_v$ that describes the clinical involvement of a LNL based on imaging: $Z_v = 0$ (`False`) indicates that the LNL $v$ is healthy based on clinical diagnosis, and $Z_v = 1$ (`True`) indicates that LNL $v$ was classified as involved. $X_v$ and $Z_v$ are connected through the sensitivity and specificity of the diagnositc modality.

@@

The transition matrix $\mathbf{A}$ is parameterized using a directed acyclic graph (DAG) that represents the underlying lymphatic network. Edges from the primary tumor to an LNL are associated with a probability $b_v$ for direct spread to LNL $v$ during one time step. Arcs from a LNL $v$ to a LNL $r$ are parameterized with the probability rate $t_{vr}$ representing the probability of spread to an LNL $r$ that receives efferent lymphatic spread from LNL $v$. In this paper, we build on the DAG shown in @fig-full-graph which was obtained by maximizing the model evidence as described in [@ludwig_modelling_2024].

@@

$$
P\left( \mathbf{X} \mid \text{T}x = \text{early} \right) = \sum_{t=0}^{t_\text{max}} P \left( \mathbf{X} \mid t \right) \cdot P(t \mid \text{T}x = \text{early})
$$

@@

The probabiltiy for a patient to present with a diagnosis $\boldsymbol{\zeta}_i$ a T-category $\text{T}x$ tumor can now be written as:

@@

which represents a single patient's contribution to the overall likelihood function.

Above, $\left[ \ldots \right]_i$ we denote the $i$-th component of the vector in the square brackets. Note that it is also possible to account for missing involvement information: If a diagnosis (like fine needle aspiration (FNA)) is only available for a subset of all LNLs, we can sum over all those possible complete observed states $\boldsymbol{\zeta}_j$ that match the provided diagnosis.

@@

<!-- This is displayed in @tbl-data-strat and shows how often the contralateral LNLs I, II, III, and IV were involved, given all possible combinations of midline extension, T-category, and ipsilateral LNL III involvement. Unsurprisingly, the prevalence for contralateral involvement is consistently higher when the tumor extends over the mid-sagittal plane or is of later T-category. But it is also more frequent when the ipsilateral side shows more severe involvement, which is here shown via the surrogate LNL III. -->

@@

Thus, we extend the formalism in @sec-unilateral in such a way that the model's ipsi- and contralateral side evolve synchronously over time. To achieve that, we start by writing down the posterior distribution of involvement, which is now a joint probability of an involvement $\mathbf{X}^\text{i}$ ipsilaterally *and* an involvement $\mathbf{X}^\text{c}$ contralaterally, given a diagnosis of the ipsilateral LNLs $\mathbf{Z}^\text{i}$ and of the contralateral ones $\mathbf{Z}^\text{c}$:

@@

The matrices $\boldsymbol{\Lambda}_\text{i}$ and $\boldsymbol{\Lambda}_\text{c}$ could, in principle, be parameterized with entirely separate parameters, allowing ipsilateral and contralateral spread rates to differ completely. However, we simplify the parameter space by sharing parameters between the two sides, based on the following three assumptions:

1. **Shared Graph Structure**: Both ipsilateral and contralateral spread are described by the same graph shown in @fig-full-graph.
2. **Symmetric Spread Among LNLs**: The spread among LNLs is assumed to be the same on both sides, reflecting the symmetric structure of the lymphatic system. Consequently, the spread rates between nodes should also be symmetric. This is formalized as:\
  $$
  t_{rv}^\text{c} = t_{rv}^\text{i}
  $$ {#eq-symmetries}
  for all $v \leq V$ and $r \in \operatorname{pa}(v)$.
3. **Asymmetric Spread from Tumor**: Tumor spread to the contralateral side varies depending on whether the tumor crosses the mid-sagittal plane. When the tumor extends over the midline, its spread rate to the contralateral LNLs should be larger than if the tumor were lateralized, but still smaller than the tumor's spread to the ipsilateral side. To model this, we introduce a mixing parameter $\alpha \in [0,1]$, representing the "degree of asymmetry" effectively mixing these two extremes:\
  $$
  b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
  $$ {#eq-mixing}
  This parameterization results in three sets of rates for tumor spread to the LNLs: (1) the spread to ipsilateral LNLs $b^\text{i}_v$, (2) the spread to contralateral LNLs as long as the tumor is lateralized $b_v^{\text{c},\epsilon=\texttt{False}}$, (3) the spread to contralateral LNLs when the tumor crosses the midline $b_v^{\text{c},\epsilon=\texttt{True}}$. Despite this additional complexity, these rates collectively account for only $2 \cdot V + 1$ parameters due to their coupling via $\alpha$.

@@

To account for this, we model the tumor's extension over the mid-sagittal plane as a binary random variable. A tumor starts as lateralized, with a finite probability $p_\epsilon$ at each time step of crossing the midline. The overall probabilities of a patient having a clearly lateralized tumor or one extending over the mid-sagittal plane after $t$ time steps are then given by

@@

The case of midline extension is more complex: we already marginalize over the exact time step when the tumor grows over the mid-sagittal plane. However, at the point of crossing, the contralateral transition matrix must switch to the increased spread rates, $b_v^{\text{c}, \epsilon=\texttt{True}}$, as defined by the linear mixing in @eq-mixing. To correctly perform this marginalization, we iteratively construct the joint distribution $P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)$.

@@

Using this, we can again write the joint distribution of ipsi- and contralateral involvement - now also for the case of mid-sagittal extension - algebraically as before in @eq-bilateral-marginal-algebra:

@@

# Computational Methods {#sec-methods}

@@

A "scenario" includes the patient's T-category $\text{T}x$ and whether the tumor extended over the mid-sagittal plane, i.e. $\epsilon=\texttt{True}$ or $\epsilon=\texttt{False}$. An involvement pattern specifies all ipsi- and contralateral LNLs' status as "healthy", "involved", or "masked" (ignored).

@@

![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences, shown for the contralateral LNLs II (blue), III (orange), and IV (green). The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones. The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal plane. This figure illustrates the model's ability to describe the prevalence of involvement for different combinations of the risk factors T-category and midline extension.](figures/fig_model_prevalences_overall.svg){#fig-model-prevalences-overall}

@@

![Comparison of the computed and observed prevalences for scenarios that illustrate the model's capability of accounting for the correlation between ipsi- and contralateral involvement. We show three scenarios where we consider the joint involvement of contralateral LNL II together with different ipsilateral involvements: 1) the ipsilateral neck shows no involvement in green (LNLs I to V are healthy, LNL VII is unspecified because data on it is missing for some patients), 2) where ipsilateral LNL II is involved in orange (LNLs I, III, IV, and V are healthy), and 3) where ipsilateral LNLs II and III are involved in red (LNLs I, IV, and V are healthy). These three scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal plane in the right column).](figures/fig_model_prevalences_with_ipsi.svg){#fig-model-prevalences-with-ipsi}

@@

```{python}
#| echo: false
#| label: tbl-data-strat
#| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal plane, the T-category, and whether the ipsilateral LNL III was involved or healthy.
lnl_cols = get_lnl_cols("contra", lnls=["I", "II", "III", "IV"])
num_ipsi_inv = raw[get_lnl_cols("ipsi")].sum(axis="columns")
